Deriving Random Generators via\\ Higher-Level Representations

### 2019-01-17 paper/main.tex
WIP

% Local Variables:
% TeX-master: "main.lhs.tex"
% TeX-command-default: "Make"
% End:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title
\title{Synthesizing Random Generators via Higher-Order Representations}

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\text{HRep}_{Val}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}\text{Mk}_{Val}\;\Conid{Int}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{Add}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}\text{Mk}_{Add}\;\Varid{r}\;\Varid{r}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{Mul}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}\text{Mk}_{Mul}\;\Varid{r}\;\Varid{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Note that the previous definitions are type parametric over the type parameter
\ensuremath{\Varid{r}}.
%
This allow us to replace \ensuremath{\Varid{r}} with any concrete data type, obtaining different
possible values on each case.
%
For instance, the value (\ensuremath{\text{Mk}_{Add}\;\mathrm{10}\;\mathrm{20}}) has type \ensuremath{\text{HRep}_{Add}\;\Conid{Int}}, while the
value (\ensuremath{\text{Mk}_{Mul}\;\Conid{True}\;\Conid{False}}) has type \ensuremath{\text{HRep}_{Mul}\;\Conid{Bool}}.
%
In practice, this parametricity let us instatiate \ensuremath{\Varid{r}} with the type of the
chosen generation specification (which might be composed of several \ensuremath{\text{HRep}}s),
without having to modify anything in the underlying machinery.

Having the \ensuremath{\text{HRep}} of each data constructor, we can define an evaluation relation
($\evalrep{\_}{t} : \ensuremath{\text{HRep}}_f \rightarrow t$) that maps a value from each
representation \ensuremath{\Varid{f}} back to the target data type \ensuremath{\Varid{t}}.
%
Then, we simply need to translate each constructor representation back into its
corresponding one, translating the abstracted fields recursively:
%
\begin{alignat*}{4}
  &\evalrep{\ensuremath{\text{Mk}_{Val}}\ n\ \ \ &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Val}}\ n \\
  &\evalrep{\ensuremath{\text{Mk}_{Add}}\ x\ y   &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Add}}\ \evalrep{x}{Exp}\ \evalrep{y}{Exp} \\
  &\evalrep{\ensuremath{\text{Mk}_{Mul}}\ x\ y   &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Mul}}\ \evalrep{x}{Exp}\ \evalrep{y}{Exp}
\end{alignat*}

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}gen_{Val}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}\text{Mk}_{Val}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}{}\<[E]%
\\
\>[3]{}gen_{Add}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}\text{Mk}_{Add}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\\
\>[3]{}gen_{Mul}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}\text{Mk}_{Mul}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

@@

In this work we define type level combinators which let the user combine the
automatically derived \ensuremath{\text{HRep}}s in several ways.
%
In first place, we define a type combinator ($\_^\bullet$) to tag a \ensuremath{\text{HRep}} to be
terminal, i.e., a representation that is allowed be generated when the
generation size gets exausted:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;(f^\bullet)\;\Varid{a}\mathrel{=}\Conid{Term}\;(\Varid{f}\;\Varid{a}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

The previous combinators only include information relevant to the generation
process, in a sense that neither one adds new structure to the final
representation.
%
In this light, they do not alter the evaluation semantics, and we translate them
back to our target data type by evaluating the inner representation:
%
\begin{alignat*}{4}
  &\evalrep{Freq\ x &&: f \otimes n &&}{t} &&= \evalrep{x : f}{t} \\
  &\evalrep{Term\ x &&: \ensuremath{f^\bullet}    &&}{t} &&= \evalrep{x : f}{t}
\end{alignat*}

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{31}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}gen_{f^\bullet}\;{}\<[16]%
\>[16]{}gen_{r}{}\<[23]%
\>[23]{}\mathrel{=}\Conid{Term}{}\<[31]%
\>[31]{}\mathop{\mathsmaller{\langle \$ \rangle}}gen_{f}\;gen_{r}{}\<[E]%
\\
\>[B]{}gen_{f \otimes n}\;{}\<[16]%
\>[16]{}gen_{r}{}\<[23]%
\>[23]{}\mathrel{=}\Conid{Freq}{}\<[31]%
\>[31]{}\mathop{\mathsmaller{\langle \$ \rangle}}gen_{f}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

With the introduced combinators, we can easily create a type synonym \ensuremath{\text{HRep}_{Exp}}
to specify a generation schema equivalent to the one seen in the concrete random
generator of type \ensuremath{\Conid{Exp}} presented in Section \ref{sec:randomtesting}:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\text{HRep}_{Exp}{}\<[16]%
\>[16]{}\mathrel{=}{}\<[24]%
\>[24]{}\text{HRep}_{Val}^\bullet{}\<[E]%
\\
\>[16]{}\!\;\oplus\;{}\<[24]%
\>[24]{}\text{HRep}_{Add}\;{}\<[34]%
\>[34]{}\otimes\;\mathrm{2}{}\<[E]%
\\
\>[16]{}\!\;\oplus\;{}\<[24]%
\>[24]{}\text{HRep}_{Mul}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

However, a value of type \ensuremath{\text{HRep}_{Exp}} still has its recursive calls abstracted
away---the type parameter \ensuremath{\Varid{r}} is implicit in the definition of \ensuremath{\text{HRep}_{Exp}}.
%
If we want to generate recursive values using this representation we need to
introduce a last type level combinator to ``tie the knot'' recursively:

@@

Unlike the other combinators, to randomly generate a fixed point of a certain
representation \ensuremath{\Varid{f}}, we do not parametrize the generation of the recursive fields
of the inner represention over an external generator \ensuremath{gen_{r}}.
%
Instead, we can also tie the recursive knot in our random generator by piping
the generation of the inner \ensuremath{\text{HRep}} parametrized recursively by the itself:

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}gen_{Exp}\mathrel{=}\mathbf{do}\;{}\<[17]%
\>[17]{}\Varid{x}\leftarrow genFix_{(\text{HRep}\,{Exp})}{}\<[E]%
\\
\>[17]{}\Varid{return}\;\evalrep{x}{Exp}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\text{HRep}_{foo\#1}\;{}\<[18]%
\>[18]{}\Varid{r}\mathrel{=}\text{Mk}_{foo\#1}\;\Varid{r}\;\Varid{r}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{foo\#2}\;{}\<[18]%
\>[18]{}\Varid{r}\mathrel{=}\text{Mk}_{foo\#2}\;\Conid{Int}\;\Varid{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\begin{align*}
  &\evalrep{\ensuremath{\text{Mk}_{foo\#1}}\ x\ y}{Exp} = \\
  &\ \ \ensuremath{\Conid{Add}\;(\Conid{Add}\;\evalrep{x}{Exp}\;(\Conid{Val}\;\mathrm{50}))\;(\Conid{Add}\;(\Conid{Val}\;\mathrm{25})\;\evalrep{y}{Exp})}\\
  &\evalrep{\ensuremath{\text{Mk}_{foo\#2}}\ x\ y}{Exp} = \\
  &\ \ \ensuremath{\Conid{Mul}\;(\Conid{Val}\;\mathrm{50})\;(\Conid{Mul}\;(\Conid{Val}\;\Varid{x})\;\evalrep{y}{Exp})}
\end{align*}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
\column{52}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}gen_{foo\#1}\;{}\<[14]%
\>[14]{}gen_{r}{}\<[21]%
\>[21]{}\mathrel{=}\text{Mk}_{foo\#1}{}\<[33]%
\>[33]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}{}\<[52]%
\>[52]{}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\\
\>[3]{}gen_{foo\#2}\;{}\<[14]%
\>[14]{}gen_{r}{}\<[21]%
\>[21]{}\mathrel{=}\text{Mk}_{foo\#2}{}\<[33]%
\>[33]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}{}\<[52]%
\>[52]{}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{38}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\text{HRep}_{foo}{}\<[16]%
\>[16]{}\mathrel{=}\text{HRep}_{foo\#1}\;\!\!\oplus\!\!\;{}\<[38]%
\>[38]{}\text{HRep}_{foo\#2}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\text{HRep}_{ten}\;{}\<[21]%
\>[21]{}\Varid{r}\mathrel{=}\text{Mk}_{ten}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{square}\;{}\<[21]%
\>[21]{}\Varid{r}\mathrel{=}\text{Mk}_{square}\;{}\<[37]%
\>[37]{}\Varid{r}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\text{HRep}_{minus}\;{}\<[21]%
\>[21]{}\Varid{r}\mathrel{=}\text{Mk}_{minus}\;{}\<[37]%
\>[37]{}\Varid{r}\;\Varid{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\begin{alignat*}{3}
  &\evalrep{\ensuremath{\text{Mk}_{ten}}         &&}{Exp} = \ensuremath{\Varid{ten}} \\
  &\evalrep{\ensuremath{\text{Mk}_{square}}\ x   &&}{Exp} = \ensuremath{\Varid{square}\;\evalrep{x}{Exp}} \\
  &\evalrep{\ensuremath{\text{Mk}_{minus}}\ x\ y &&}{Exp} = \ensuremath{\Varid{minus}\;\evalrep{x}{Exp}\;\evalrep{y}{Exp}}
\end{alignat*}

>>>>

% Local Variables:
% TeX-mbigstarer: "main.lhs.tex"
% TeX-command-default: "Make"
% End:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title
\title{Deriving Random Generators via\\ Higher-Level Representations}

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;{H\!Rep}_{Val}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}{Mk}_{Val}\;\Conid{Int}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;{H\!Rep}_{Add}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}{Mk}_{Add}\;\Varid{r}\;\Varid{r}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;{H\!Rep}_{Mul}\;{}\<[16]%
\>[16]{}\Varid{r}\mathrel{=}{Mk}_{Mul}\;\Varid{r}\;\Varid{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Note that the previous definitions are type parametric over the type parameter
\ensuremath{\Varid{r}}.
%
This allow us to replace \ensuremath{\Varid{r}} with any concrete data type, obtaining different
possible values on each case.
%
For instance, the value (\ensuremath{{Mk}_{Add}\;\mathrm{10}\;\mathrm{20}}) has type \ensuremath{{H\!Rep}_{Add}\;\Conid{Int}}, while the
value (\ensuremath{{Mk}_{Mul}\;\Conid{True}\;\Conid{False}}) has type \ensuremath{{H\!Rep}_{Mul}\;\Conid{Bool}}.
%
In practice, this parametricity let us instatiate \ensuremath{\Varid{r}} with the type of the
chosen generation specification (which might be composed of several \ensuremath{\text{HRep}}s),
without having to modify anything in the underlying machinery.

Having the \ensuremath{\text{HRep}} of each data constructor, we can define an evaluation relation
($\evalrep{\_}{t} : \ensuremath{\text{HRep}}_f \rightarrow t$) that maps a value from each
representation \ensuremath{\Varid{f}} back to the target data type \ensuremath{\Varid{t}}.
%
Then, we simply need to translate each constructor representation back into its
corresponding one, translating the abstracted fields recursively:
%
\begin{alignat*}{4}
  &\evalrep{\ensuremath{{Mk}_{Val}}\ n\ \ \ &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Val}}\ n \\
  &\evalrep{\ensuremath{{Mk}_{Add}}\ x\ y   &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Add}}\ \evalrep{x}{Exp}\ \evalrep{y}{Exp} \\
  &\evalrep{\ensuremath{{Mk}_{Mul}}\ x\ y   &&}{\ensuremath{\Conid{Exp}}}
    &&= \ensuremath{\Conid{Mul}}\ \evalrep{x}{Exp}\ \evalrep{y}{Exp}
\end{alignat*}

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}gen_{Val}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}{Mk}_{Val}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}{}\<[E]%
\\
\>[3]{}gen_{Add}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}{Mk}_{Add}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\\
\>[3]{}gen_{Mul}\;{}\<[12]%
\>[12]{}gen_{r}{}\<[19]%
\>[19]{}\mathrel{=}{Mk}_{Mul}{}\<[29]%
\>[29]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

@@

In this work we define type level combinators which let the user combine the
automatically derived \ensuremath{\text{HRep}}s in several ways.
%
In first place, we define a type combinator (\ensuremath{\_^\bigstar}) to tag a \ensuremath{\text{HRep}} to be
terminal, i.e., a representation that is allowed be generated when the
generation size gets exausted:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;(f^\bigstar)\;\Varid{a}\mathrel{=}\Conid{Term}\;(\Varid{f}\;\Varid{a}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

The previous combinators only include information relevant to the generation
process, in a sense that neither one adds new structure to the final
representation.
%
In this light, they do not alter the evaluation semantics, and we translate them
back to our target data type by evaluating the inner representation:
%
\begin{alignat*}{4}
  &\evalrep{Freq\ x &&: f \otimes n &&}{t} &&= \evalrep{x : f}{t} \\
  &\evalrep{Term\ x &&: \ensuremath{f^\bigstar}    &&}{t} &&= \evalrep{x : f}{t}
\end{alignat*}

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{31}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}gen_{f^\bigstar}\;{}\<[16]%
\>[16]{}gen_{r}{}\<[23]%
\>[23]{}\mathrel{=}\Conid{Term}{}\<[31]%
\>[31]{}\mathop{\mathsmaller{\langle \$ \rangle}}gen_{f}\;gen_{r}{}\<[E]%
\\
\>[B]{}gen_{f \otimes n}\;{}\<[16]%
\>[16]{}gen_{r}{}\<[23]%
\>[23]{}\mathrel{=}\Conid{Freq}{}\<[31]%
\>[31]{}\mathop{\mathsmaller{\langle \$ \rangle}}gen_{f}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

With the introduced combinators, we can easily create a type synonym \ensuremath{{H\!Rep}_{Exp}}
to specify a generation schema equivalent to the one seen in the concrete random
generator of type \ensuremath{\Conid{Exp}} presented in Section \ref{sec:randomtesting}:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;{H\!Rep}_{Exp}{}\<[16]%
\>[16]{}\mathrel{=}{}\<[24]%
\>[24]{}{H\!Rep}_{Val}^\bigstar{}\<[E]%
\\
\>[16]{}\!\;\oplus\;{}\<[24]%
\>[24]{}{H\!Rep}_{Add}\;{}\<[34]%
\>[34]{}\otimes\;\mathrm{2}{}\<[E]%
\\
\>[16]{}\!\;\oplus\;{}\<[24]%
\>[24]{}{H\!Rep}_{Mul}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

However, a value of type \ensuremath{{H\!Rep}_{Exp}} still has its recursive calls abstracted
away---the type parameter \ensuremath{\Varid{r}} is implicit in the definition of \ensuremath{{H\!Rep}_{Exp}}.
%
If we want to generate recursive values using this representation we need to
introduce a last type level combinator to ``tie the knot'' recursively:

@@

Unlike the other combinators, to randomly generate a fixed point of a certain
representation \ensuremath{\Varid{f}}, we do not parametrize the generation of the recursive fields
of the inner represention over an external generator \ensuremath{gen_{r}}.
%
Instead, we can tie the recursive knot in our random generator by piping the
generation of the inner \ensuremath{\text{HRep}}, and calling itself to generate recursive values:

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}gen_{Exp}\mathrel{=}\mathbf{do}\;{}\<[17]%
\>[17]{}\Varid{x}\leftarrow genFix_{\,\text{HRep}_{Exp}}{}\<[E]%
\\
\>[17]{}\Varid{return}\;\evalrep{x}{Exp}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

So far we have introduced the representation machinery required to represent the
random generation of a target data type considering only the structure encoded
on its definition.
%
This approach has the advantage of being extendable with different sources of
structured information.
%
We proceed to introduce two extensions that help to address the problematic
testing scenarios presented in the previous section.

@@

We can follow a similar reasoning as before to represent the pattern matching
structure from a given function.
%
Consider the function \ensuremath{\Varid{foo}} defined in the previous section.
%
We can derive a new data type for each one of its pattern matchings, whose
fields represent each pattern variable present on the given pattern, abstracting
away every pattern variable of type \ensuremath{\Conid{Exp}}.
%
Concretely, we define the following data types:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;{H\!Rep}_{foo\#1}\;{}\<[18]%
\>[18]{}\Varid{r}\mathrel{=}{Mk}_{foo\#1}\;\Varid{r}\;\Varid{r}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;{H\!Rep}_{foo\#2}\;{}\<[18]%
\>[18]{}\Varid{r}\mathrel{=}{Mk}_{foo\#2}\;\Conid{Int}\;\Varid{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The first pattern of \ensuremath{\Varid{foo}} conteins two pattern variables (\ensuremath{\Varid{x}} and \ensuremath{\Varid{y}}) of type
\ensuremath{\Conid{Exp}} that are abstracted in the fields of \ensuremath{{Mk}_{foo\#1}}.
%
Similarly, the second pattern of \ensuremath{\Varid{foo}} contains the pattern variable \ensuremath{\Varid{x}} of type
\ensuremath{\Conid{Int}} represented by the first field of \ensuremath{{Mk}_{foo\#2}}, along with the pattern
variable \ensuremath{\Varid{y}} of type \ensuremath{\Conid{Exp}} that also is abstracted in the second field of
\ensuremath{{Mk}_{foo\#2}}.

To evaluate these representations back to our target data type, we simply expand
them it into the concrete value represented by the original pattern:
%
\begin{align*}
  &\evalrep{\ensuremath{{Mk}_{foo\#1}}\ x\ y}{Exp} = \\
  &\ \ \ensuremath{\Conid{Add}\;(\Conid{Add}\;\evalrep{x}{Exp}\;(\Conid{Val}\;\mathrm{50}))\;(\Conid{Add}\;(\Conid{Val}\;\mathrm{25})\;\evalrep{y}{Exp})}\\
  &\evalrep{\ensuremath{{Mk}_{foo\#2}}\ x\ y}{Exp} = \\
  &\ \ \ensuremath{\Conid{Mul}\;(\Conid{Val}\;\mathrm{50})\;(\Conid{Mul}\;(\Conid{Val}\;\Varid{x})\;\evalrep{y}{Exp})}
\end{align*}

And the random generation of pattern \ensuremath{\text{HRep}}s is defined in the same way as we
did before with data constructors \ensuremath{\text{HRep}}s:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
\column{52}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}gen_{foo\#1}\;{}\<[14]%
\>[14]{}gen_{r}{}\<[21]%
\>[21]{}\mathrel{=}{Mk}_{foo\#1}{}\<[33]%
\>[33]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}{}\<[52]%
\>[52]{}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\\
\>[3]{}gen_{foo\#2}\;{}\<[14]%
\>[14]{}gen_{r}{}\<[21]%
\>[21]{}\mathrel{=}{Mk}_{foo\#2}{}\<[33]%
\>[33]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}{}\<[52]%
\>[52]{}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

Finally, we can join the pattern matching representation of each clause of \ensuremath{\Varid{foo}}
into a single one:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{38}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;{H\!Rep}_{foo}{}\<[16]%
\>[16]{}\mathrel{=}{H\!Rep}_{foo\#1}\;\!\!\oplus\!\!\;{}\<[38]%
\>[38]{}{H\!Rep}_{foo\#2}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

To introduce the higher level representation of module abstract interfaces,
suppose we have a module \ensuremath{\Conid{M}} defining \ensuremath{\Conid{Exp}} combinators as follows:

@@

We will represent each function of \ensuremath{\Conid{M}} returning a value of type \ensuremath{\Conid{Exp}} following
the same idea as before, deriving a data type with a single data constructor for
each one of them:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;{H\!Rep}_{ten}\;{}\<[21]%
\>[21]{}\Varid{r}\mathrel{=}{Mk}_{ten}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;{H\!Rep}_{square}\;{}\<[21]%
\>[21]{}\Varid{r}\mathrel{=}{Mk}_{square}\;{}\<[37]%
\>[37]{}\Varid{r}{}\<[E]%
\\
\>[B]{}\mathbf{data}\;{H\!Rep}_{minus}\;{}\<[21]%
\>[21]{}\Varid{r}\mathrel{=}{Mk}_{minus}\;{}\<[37]%
\>[37]{}\Varid{r}\;\Varid{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

In this case, each single constructor will have as fields the types of the
inputs of the function that they represent.
%
As before, we abstract away any field representing and input of type \ensuremath{\Conid{Exp}} with
a type parameter \ensuremath{\Varid{r}}.

To evaluate these representations, we simply pipe the values on its fields as
input parameter of each original function, returning its result.

\begin{alignat*}{3}
  &\evalrep{\ensuremath{{Mk}_{ten}}         &&}{Exp} = \ensuremath{\Varid{ten}} \\
  &\evalrep{\ensuremath{{Mk}_{square}}\ x   &&}{Exp} = \ensuremath{\Varid{square}\;\evalrep{x}{Exp}} \\
  &\evalrep{\ensuremath{{Mk}_{minus}}\ x\ y &&}{Exp} = \ensuremath{\Varid{minus}\;\evalrep{x}{Exp}\;\evalrep{y}{Exp}}
\end{alignat*}

Moreover, the generation procedure for abstract interface representations
follows the same pattern as before:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{54}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}gen_{ten}\;{}\<[15]%
\>[15]{}gen_{r}{}\<[22]%
\>[22]{}\mathrel{=}{Mk}_{ten}{}\<[E]%
\\
\>[3]{}gen_{square}\;{}\<[15]%
\>[15]{}gen_{r}{}\<[22]%
\>[22]{}\mathrel{=}{Mk}_{square}{}\<[35]%
\>[35]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\\
\>[3]{}gen_{minus}\;{}\<[14]%
\>[14]{}gen_{r}{}\<[22]%
\>[22]{}\mathrel{=}{Mk}_{minus}{}\<[35]%
\>[35]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{r}{}\<[54]%
\>[54]{}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

We will also define a type synonym to join all the representations of the module
\ensuremath{\Conid{M}} into a single one.
